---
title: "Treatment Response and Hard Drug Use in HIV-Infected Men on HAART"
author: "Charlotte_Yunjing_Hong"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6,
  fig.align  = "center"
)
```

# Introduction

This analysis uses data from the Multicenter AIDS Cohort Study (MACS), a prospective cohort study
tracking HIV-1 infection among homosexual and bisexual men in four U.S. cities. The goal is to
determine whether treatment response 2 years after starting highly active antiretroviral therapy
(HAART) differs between men who reported hard drug use (heroin, opiates, or injection drugs) at
baseline versus those who did not.

Laboratory and animal studies have raised the possibility that hard drugs suppress immune function
and promote viral replication, but human evidence has been inconsistent. Understanding this
relationship matters clinically: if hard drug users fare worse on treatment, providers may need to
pursue more intensive HIV management or actively facilitate access to substance use treatment.

Four outcomes are examined at Year 2: viral load (VLOAD), CD4+ T cell count (LEU3N), physical
quality of life (AGG_PHYS), and mental quality of life (AGG_MENT). Because treatment adherence
may differ by drug use status and could explain any observed differences in outcomes, we also
examine whether adherence mediates the association between hard drug use and treatment response.
Both frequentist and Bayesian regression frameworks are applied and compared.

---

# Methods

## Data Cleaning

```{r libraries}
library(tidyverse)
library(tableone)
library(knitr)
library(kableExtra)
library(ggplot2)
library(patchwork)
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(HDInterval)
```

```{r load-data}
dat_raw <- read.csv("/mnt/user-data/uploads/hiv_6624_final__3_.csv",
                    stringsAsFactors = FALSE)
```

```{r data-cleaning}
# -------------------------------------------------------------------
# Step 1: Recode special/sentinel values to NA
# -------------------------------------------------------------------
dat <- dat_raw %>%
  mutate(
    # Recode improbable / insufficient values
    AGG_PHYS = ifelse(AGG_PHYS %in% c(-1, 9, 99, 999), NA, AGG_PHYS),
    AGG_MENT = ifelse(AGG_MENT %in% c(-1, 9, 99, 999), NA, AGG_MENT),
    VLOAD    = ifelse(VLOAD    %in% c(-1, 9, 99, 999999999), NA, VLOAD),
    LEU3N    = ifelse(LEU3N    %in% c(-1), NA, LEU3N),
    CESD     = ifelse(CESD     == -1, NA, CESD),
    BMI      = ifelse(BMI < 10 | BMI > 70, NA, BMI),
    ADH      = ifelse(ADH %in% c(NA, 9), NA, ADH),   # adherence
    income   = ifelse(income == 9, NA, income),
    EDUCBAS  = ifelse(EDUCBAS %in% c(NA, 9), NA, EDUCBAS),
    DKGRP    = ifelse(DKGRP %in% c(NA, 9), NA, DKGRP),
    SMOKE    = ifelse(SMOKE %in% c(NA, 9), NA, SMOKE)
  )

# -------------------------------------------------------------------
# Step 2: Log10-transform viral load (no zeros in VLOAD based on codebook)
# -------------------------------------------------------------------
# Check for zeros
# table(dat$VLOAD == 0, useNA = "always")  # uncomment to inspect
dat <- dat %>%
  mutate(log10VLOAD = log10(VLOAD + 1))  # +1 as precaution for any zero values

# -------------------------------------------------------------------
# Step 3: Create categorical variables per professor's instructions
# -------------------------------------------------------------------
dat <- dat %>%
  mutate(
    # Race: White vs Other (2 groups, as per class)
    race_cat = factor(
      ifelse(RACE == 1, "White", "Other"),
      levels = c("White", "Other")
    ),

    # Education: College degree+ vs No college degree (2 groups)
    educ_cat = factor(
      ifelse(EDUCBAS >= 5, "College degree+", "No college degree"),
      levels = c("No college degree", "College degree+")
    ),

    # Smoking: 3 levels
    smoke_cat = factor(
      case_when(
        SMOKE == 1 ~ "Never",
        SMOKE == 2 ~ "Former",
        SMOKE == 3 ~ "Current",
        TRUE       ~ NA_character_
      ),
      levels = c("Never", "Former", "Current")
    ),

    # Adherence as ordered factor (1=100%, 2=95-99%, 3=75-94%, 4=<75%)
    adh_cat = factor(ADH, levels = 1:4,
                     labels = c("100%", "95-99%", "75-94%", "<75%"),
                     ordered = TRUE),
    
    # Binary adherence: perfect vs imperfect
    adh_perfect = ifelse(ADH == 1, 1, 0),

    # Hard drug use as factor
    hard_drugs_f = factor(hard_drugs, levels = c(0, 1),
                          labels = c("No", "Yes"))
  )

# -------------------------------------------------------------------
# Step 4: Identify baseline (year 0) and Year 2 records
# -------------------------------------------------------------------
bl <- dat %>% filter(years == 0, hivpos == 1) %>%
  select(newid, hard_drugs_f,
         log10VLOAD_bl = log10VLOAD,
         LEU3N_bl      = LEU3N,
         AGG_PHYS_bl   = AGG_PHYS,
         AGG_MENT_bl   = AGG_MENT,
         age_bl        = age,
         BMI_bl        = BMI,
         race_cat, educ_cat, smoke_cat,
         CESD_bl = CESD,
         DKGRP_bl = DKGRP,
         income_bl = income)

y2 <- dat %>% filter(years == 2) %>%
  select(newid,
         log10VLOAD_y2 = log10VLOAD,
         LEU3N_y2      = LEU3N,
         AGG_PHYS_y2   = AGG_PHYS,
         AGG_MENT_y2   = AGG_MENT,
         ADH_y2        = ADH,
         adh_perfect_y2 = adh_perfect)

# -------------------------------------------------------------------
# Step 5: Merge and restrict to analytic sample
# Consistent set of subjects with all 4 outcomes (per class discussion)
# -------------------------------------------------------------------
analytic_full <- inner_join(bl, y2, by = "newid") %>%
  filter(!is.na(hard_drugs_f))

# Analytic sample: complete cases on all 4 outcomes + core covariates
analytic <- analytic_full %>%
  filter(
    !is.na(log10VLOAD_y2),
    !is.na(LEU3N_y2),
    !is.na(AGG_PHYS_y2),
    !is.na(AGG_MENT_y2),
    !is.na(log10VLOAD_bl),
    !is.na(LEU3N_bl),
    !is.na(AGG_PHYS_bl),
    !is.na(AGG_MENT_bl),
    !is.na(age_bl),
    !is.na(race_cat),
    !is.na(BMI_bl),
    !is.na(educ_cat),
    !is.na(smoke_cat)
  )

cat("Final analytic sample size:", nrow(analytic), "\n")
cat("Hard drug users:", sum(analytic$hard_drugs_f == "Yes"), "\n")
cat("Non-users:", sum(analytic$hard_drugs_f == "No"), "\n")
```

---

# Descriptive Statistics

## Missing Data Summary (Table A1)

```{r missing-data}
vars_check <- c("log10VLOAD_y2","LEU3N_y2","AGG_PHYS_y2","AGG_MENT_y2",
                "log10VLOAD_bl","LEU3N_bl","AGG_PHYS_bl","AGG_MENT_bl",
                "age_bl","BMI_bl","race_cat","educ_cat","smoke_cat",
                "CESD_bl","ADH_y2")

miss_tab <- data.frame(
  Variable   = vars_check,
  N_Missing  = sapply(vars_check, function(v) sum(is.na(analytic_full[[v]]))),
  Pct_Missing = sapply(vars_check, function(v)
    round(100 * mean(is.na(analytic_full[[v]])), 1))
)

kable(miss_tab, row.names = FALSE,
      caption = "Table A1: Missing Data Summary (pre-analytic sample)") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Baseline Characteristics by Hard Drug Use (Table A2)

```{r table1}
tab_vars <- c("age_bl","race_cat","educ_cat","BMI_bl","smoke_cat",
              "CESD_bl","DKGRP_bl","log10VLOAD_bl","LEU3N_bl",
              "AGG_PHYS_bl","AGG_MENT_bl")
cat_vars  <- c("race_cat","educ_cat","smoke_cat")
non_norm  <- c("CESD_bl","log10VLOAD_bl","LEU3N_bl","DKGRP_bl")

tab1 <- CreateTableOne(vars = tab_vars, strata = "hard_drugs_f",
                       data = analytic, factorVars = cat_vars,
                       addOverall = TRUE)

print(tab1, nonnormal = non_norm, showAllLevels = TRUE,
      caption = "Table A2: Baseline Characteristics by Hard Drug Use Status")
```

## Longitudinal Outcome Trajectories (Figure A1)

```{r trajectories, fig.height=8}
# Use full longitudinal data for trajectory plot
traj <- dat %>%
  filter(hivpos == 1, years %in% 0:8) %>%
  inner_join(bl %>% select(newid, hard_drugs_f), by = "newid") %>%
  group_by(years, hard_drugs_f) %>%
  summarise(
    mean_VL   = mean(log10VLOAD, na.rm = TRUE),
    mean_CD4  = mean(LEU3N,      na.rm = TRUE),
    mean_PHYS = mean(AGG_PHYS,   na.rm = TRUE),
    mean_MENT = mean(AGG_MENT,   na.rm = TRUE),
    .groups = "drop"
  ) %>% filter(!is.na(hard_drugs_f))

p1 <- ggplot(traj, aes(years, mean_VL, color = hard_drugs_f, group = hard_drugs_f)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Log10 Viral Load", x = "Years since HAART", y = "Mean Log10 VL",
       color = "Baseline\nHard Drug Use") +
  theme_bw()

p2 <- ggplot(traj, aes(years, mean_CD4, color = hard_drugs_f, group = hard_drugs_f)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "CD4+ T Cell Count", x = "Years since HAART", y = "Mean CD4 (cells/mL)",
       color = "Baseline\nHard Drug Use") +
  theme_bw()

p3 <- ggplot(traj, aes(years, mean_PHYS, color = hard_drugs_f, group = hard_drugs_f)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Physical QoL (SF-36)", x = "Years since HAART", y = "Mean Score",
       color = "Baseline\nHard Drug Use") +
  theme_bw()

p4 <- ggplot(traj, aes(years, mean_MENT, color = hard_drugs_f, group = hard_drugs_f)) +
  geom_line(size = 1) + geom_point(size = 2) +
  labs(title = "Mental QoL (SF-36)", x = "Years since HAART", y = "Mean Score",
       color = "Baseline\nHard Drug Use") +
  theme_bw()

(p1 + p2) / (p3 + p4) +
  plot_annotation(title = "Figure A1: Mean Treatment Response Trajectories by Baseline Hard Drug Use")
```

## Year 2 Outcome Distributions (Figure A3)

```{r outcome-distributions, fig.height=8}
long_y2 <- analytic %>%
  select(hard_drugs_f, log10VLOAD_y2, LEU3N_y2, AGG_PHYS_y2, AGG_MENT_y2) %>%
  pivot_longer(cols = -hard_drugs_f, names_to = "outcome", values_to = "value") %>%
  mutate(outcome = recode(outcome,
    "log10VLOAD_y2" = "Log10 Viral Load",
    "LEU3N_y2"      = "CD4+ T Cell Count",
    "AGG_PHYS_y2"   = "Physical QoL",
    "AGG_MENT_y2"   = "Mental QoL"))

ggplot(long_y2, aes(x = value, fill = hard_drugs_f)) +
  geom_histogram(alpha = 0.7, bins = 30, position = "identity") +
  facet_wrap(~ outcome, scales = "free") +
  scale_fill_manual(values = c("No" = "#F4A582", "Yes" = "#4DAC26")) +
  labs(title = "Figure A3: Year 2 Outcome Distributions by Baseline Hard Drug Use",
       x = "Value", fill = "Baseline Hard\nDrug Use") +
  theme_bw()
```

---

# Frequentist Analysis

All four models adjust for: age at baseline, race (White vs Other), BMI, education, smoking, and the baseline value of the respective outcome — consistent with the investigator's request for face validity covariates.

Additional covariates (income, alcohol use, depression [CESD], ART adherence) are evaluated for inclusion if p < 0.20 in univariable association with the outcome OR if they shift the hard drug coefficient by more than 10%.

## Covariate Selection

```{r covariate-selection}
# Function to check potential covariates
check_covariates <- function(outcome_var, df) {
  potential <- c("income_bl", "DKGRP_bl", "CESD_bl", "ADH_y2")
  results <- lapply(potential, function(cov) {
    if (all(is.na(df[[cov]]))) return(NULL)
    tmp <- df[!is.na(df[[outcome_var]]) & !is.na(df[[cov]]), ]
    fit <- lm(as.formula(paste(outcome_var, "~", cov)), data = tmp)
    p <- summary(fit)$coefficients[2, 4]
    data.frame(covariate = cov, p_value = round(p, 4))
  })
  bind_rows(Filter(Negate(is.null), results))
}

cat("--- Log10 Viral Load ---\n")
print(check_covariates("log10VLOAD_y2", analytic))

cat("--- CD4 Count ---\n")
print(check_covariates("LEU3N_y2", analytic))

cat("--- Physical QoL ---\n")
print(check_covariates("AGG_PHYS_y2", analytic))

cat("--- Mental QoL ---\n")
print(check_covariates("AGG_MENT_y2", analytic))
```

## Multivariable Regression Models (Table A3)

```{r freq-models}
# Core covariates (always included per investigator + face validity)
core_covs <- "hard_drugs_f + age_bl + race_cat + BMI_bl + educ_cat + smoke_cat"

# Build models — add CESD and ADH if selected above (p < 0.20)
# Adjust the formula below based on covariate selection results

fit_vl <- lm(
  log10VLOAD_y2 ~ hard_drugs_f + log10VLOAD_bl + age_bl + race_cat +
    BMI_bl + educ_cat + smoke_cat + CESD_bl + ADH_y2,
  data = analytic
)

fit_cd4 <- lm(
  LEU3N_y2 ~ hard_drugs_f + LEU3N_bl + age_bl + race_cat +
    BMI_bl + educ_cat + smoke_cat + CESD_bl + ADH_y2,
  data = analytic
)

fit_phys <- lm(
  AGG_PHYS_y2 ~ hard_drugs_f + AGG_PHYS_bl + age_bl + race_cat +
    BMI_bl + educ_cat + smoke_cat + CESD_bl,
  data = analytic
)

fit_ment <- lm(
  AGG_MENT_y2 ~ hard_drugs_f + AGG_MENT_bl + age_bl + race_cat +
    BMI_bl + educ_cat + smoke_cat + CESD_bl,
  data = analytic
)

# Extract hard drug coefficients
extract_hd_coef <- function(fit, outcome_label) {
  cf <- summary(fit)$coefficients
  row <- cf["hard_drugs_fYes", ]
  ci  <- confint(fit)["hard_drugs_fYes", ]
  data.frame(
    Outcome  = outcome_label,
    Estimate = round(row[1], 4),
    SE       = round(row[2], 4),
    p_value  = round(row[4], 4),
    CI_lower = round(ci[1], 4),
    CI_upper = round(ci[2], 4)
  )
}

freq_results <- bind_rows(
  extract_hd_coef(fit_vl,   "Log10 Viral Load"),
  extract_hd_coef(fit_cd4,  "CD4+ T Cell Count"),
  extract_hd_coef(fit_phys, "Physical QoL"),
  extract_hd_coef(fit_ment, "Mental QoL")
)

kable(freq_results, caption = "Table A3: Adjusted Effect of Baseline Hard Drug Use on Year 2 Outcomes (Frequentist)",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Adjusted Coefficient" = 5))
```

## Model Diagnostics (Figure A2)

```{r diagnostics, fig.height=10}
par(mfrow = c(4, 2))

# Viral Load
plot(fit_vl,  which = 1, main = "Log10 Viral Load: Residuals vs Fitted")
plot(fit_vl,  which = 2, main = "Log10 Viral Load: Q-Q Plot")

# CD4
plot(fit_cd4, which = 1, main = "CD4 Count: Residuals vs Fitted")
plot(fit_cd4, which = 2, main = "CD4 Count: Q-Q Plot")

# Physical QoL
plot(fit_phys, which = 1, main = "Physical QoL: Residuals vs Fitted")
plot(fit_phys, which = 2, main = "Physical QoL: Q-Q Plot")

# Mental QoL
plot(fit_ment, which = 1, main = "Mental QoL: Residuals vs Fitted")
plot(fit_ment, which = 2, main = "Mental QoL: Q-Q Plot")

par(mfrow = c(1, 1))
```

---

# Addressing the Role of Adherence

The investigator asked whether any differences in outcomes between hard drug users and non-users
can be explained by differences in adherence to ART. We address this question in two ways:

**1. Descriptive comparison of adherence by drug use group**

```{r adherence-descriptive}
# Adherence at Year 2 by baseline drug use
adh_tab <- analytic %>%
  filter(!is.na(ADH_y2)) %>%
  count(hard_drugs_f, ADH_y2) %>%
  group_by(hard_drugs_f) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  mutate(ADH_label = factor(ADH_y2, levels = 1:4,
                            labels = c("100%", "95-99%", "75-94%", "<75%")))

ggplot(adh_tab, aes(x = ADH_label, y = pct, fill = hard_drugs_f)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("No" = "#F4A582", "Yes" = "#4DAC26")) +
  labs(title = "Year 2 ART Adherence by Baseline Hard Drug Use",
       x = "Adherence Level", y = "Percentage (%)",
       fill = "Baseline Hard\nDrug Use") +
  theme_bw()

# Statistical comparison
chisq.test(table(analytic$hard_drugs_f, analytic$ADH_y2))
```

**2. Mediation analysis: Does adherence explain the hard drug use effect?**

We compare the hard drug use coefficient before and after adjusting for ART adherence at Year 2.
A meaningful change (>10%) suggests adherence partially explains the association.

```{r adherence-mediation}
# Function: compare HD coefficient with and without adherence
mediation_check <- function(fit_without, outcome_var, bl_var, label) {
  df_adh <- analytic %>% filter(!is.na(ADH_y2))
  
  # Re-fit without adherence on same adherence-complete sample
  f_base <- update(fit_without, data = df_adh)
  
  # Fit with adherence
  f_adh  <- update(f_base, . ~ . + ADH_y2)
  
  beta_no_adh  <- coef(f_base)["hard_drugs_fYes"]
  beta_with_adh <- coef(f_adh)["hard_drugs_fYes"]
  pct_change   <- 100 * (beta_no_adh - beta_with_adh) / abs(beta_no_adh)
  
  data.frame(
    Outcome               = label,
    Beta_NoAdherence      = round(beta_no_adh,   4),
    Beta_WithAdherence    = round(beta_with_adh,  4),
    Pct_Change            = round(pct_change,     1),
    Adherence_Explains    = ifelse(abs(pct_change) > 10, "Yes (>10% change)", "No")
  )
}

med_results <- bind_rows(
  mediation_check(fit_vl,   "log10VLOAD_y2", "log10VLOAD_bl", "Log10 Viral Load"),
  mediation_check(fit_cd4,  "LEU3N_y2",      "LEU3N_bl",      "CD4+ T Cell Count"),
  mediation_check(fit_phys, "AGG_PHYS_y2",   "AGG_PHYS_bl",   "Physical QoL"),
  mediation_check(fit_ment, "AGG_MENT_y2",   "AGG_MENT_bl",   "Mental QoL")
)

kable(med_results,
      caption = "Table: Hard Drug Use Coefficient Before and After Adjusting for ART Adherence",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

---

# Bayesian Analysis

## Stan Model

```{r stan-model}
stan_code <- '
data {
  int<lower=0> N;
  int<lower=0> K;
  matrix[N, K] X;
  vector[N] y;
  vector[K] beta_prior_mean;
  vector<lower=0>[K] beta_prior_sd;
  real<lower=0> sigma_prior_sd;
}
parameters {
  vector[K] beta;
  real<lower=0> sigma;
}
model {
  beta  ~ normal(beta_prior_mean, beta_prior_sd);
  sigma ~ normal(0, sigma_prior_sd);
  y     ~ normal(X * beta, sigma);
}
generated quantities {
  vector[N] log_lik;
  for (n in 1:N)
    log_lik[n] = normal_lpdf(y[n] | X[n] * beta, sigma);
}
'
mod <- cmdstan_model(write_stan_file(stan_code))
```

## Helper Functions

```{r bayes-helpers}
# Build design matrix for a given outcome (consistent analytic sample)
build_stan_data <- function(outcome_var, bl_var, prior_sd_hd,
                            df = analytic, vague_sd = 100) {
  df_cc <- df %>%
    filter(complete.cases(.data[[outcome_var]], .data[[bl_var]],
                          hard_drugs_f, age_bl, race_cat, BMI_bl,
                          educ_cat, smoke_cat)) %>%
    mutate(
      hd_num     = as.numeric(hard_drugs_f == "Yes"),
      age_c      = scale(age_bl)[,1],
      bl_c       = scale(.data[[bl_var]])[,1],
      bmi_c      = scale(BMI_bl)[,1],
      race_other = as.numeric(race_cat == "Other"),
      educ_hi    = as.numeric(educ_cat == "College degree+"),
      smoke_form = as.numeric(smoke_cat == "Former"),
      smoke_curr = as.numeric(smoke_cat == "Current")
    )

  X <- model.matrix(~ hd_num + bl_c + age_c + race_other +
                      bmi_c + educ_hi + smoke_form + smoke_curr,
                    data = df_cc)
  y <- df_cc[[outcome_var]]
  K <- ncol(X)

  prior_sd            <- rep(vague_sd, K)
  prior_sd[2]         <- prior_sd_hd   # hd_num is 2nd column (after intercept)

  list(
    stan_data = list(N = nrow(X), K = K, X = X, y = y,
                     beta_prior_mean = rep(0, K),
                     beta_prior_sd   = prior_sd,
                     sigma_prior_sd  = vague_sd),
    df_cc = df_cc
  )
}

# Summarize posterior for hard drug coefficient
summarize_posterior <- function(fit, gamma, outcome_label, prior_label) {
  draws <- as_draws_df(fit$draws())
  b     <- draws$`beta[2]`
  hpdi  <- hdi(b, credMass = 0.95)
  data.frame(
    Outcome       = outcome_label,
    Prior         = prior_label,
    Post_Mean     = round(mean(b), 4),
    Post_SD       = round(sd(b),   4),
    HPDI_Lower    = round(hpdi[1], 4),
    HPDI_Upper    = round(hpdi[2], 4),
    P_meaningful  = round(mean(abs(b) > gamma), 4)
  )
}
```

## Fit Bayesian Models

```{r bayesian-fits, cache=TRUE, results='hide'}
# Outcome-specific thresholds for P(|beta| > gamma | data)
gammas <- list(
  log10VLOAD_y2 = 0.5,
  LEU3N_y2      = 50,
  AGG_PHYS_y2   = 2,   # 2 points per professor's correction
  AGG_MENT_y2   = 2
)

outcomes <- list(
  list(y = "log10VLOAD_y2", bl = "log10VLOAD_bl", label = "Log10 Viral Load"),
  list(y = "LEU3N_y2",      bl = "LEU3N_bl",      label = "CD4+ T Cell Count"),
  list(y = "AGG_PHYS_y2",   bl = "AGG_PHYS_bl",   label = "Physical QoL"),
  list(y = "AGG_MENT_y2",   bl = "AGG_MENT_bl",   label = "Mental QoL")
)

bayes_results  <- list()
loo_results    <- list()
fit_store      <- list()

for (o in outcomes) {
  for (prior_type in c("Vague", "Skeptical")) {
    hd_sd <- if (prior_type == "Vague") 100 else 10
    sd_obj <- build_stan_data(o$y, o$bl, prior_sd_hd = hd_sd)

    fit <- mod$sample(
      data             = sd_obj$stan_data,
      chains           = 4,
      parallel_chains  = 4,
      iter_sampling    = 20000,
      iter_warmup      = 5000,
      adapt_delta      = 0.90,
      seed             = 42,
      refresh          = 0
    )

    key <- paste0(o$label, "_", prior_type)
    fit_store[[key]] <- fit

    bayes_results[[key]] <- summarize_posterior(
      fit, gamma = gammas[[o$y]], o$label, prior_type
    )

    # LOO-IC (use as model comparison metric; per professor's guidance)
    ll_mat <- as_draws_matrix(fit$draws("log_lik"))
    loo_results[[key]] <- loo(ll_mat)
  }
}
```

## MCMC Diagnostics (Figure A-Bayes)

```{r mcmc-diagnostics, fig.height=6}
# Illustrate diagnostics for one outcome (Log10 VL, vague prior)
fit_vl_vague <- fit_store[["Log10 Viral Load_Vague"]]
fit_vl_vague$cmdstan_diagnose()

mcmc_trace(fit_vl_vague$draws(), pars = c("beta[1]","beta[2]","sigma"),
           facet_args = list(ncol = 1)) +
  ggtitle("Trace Plots: Log10 Viral Load (Vague Prior)")

mcmc_dens_overlay(fit_vl_vague$draws(), pars = c("beta[1]","beta[2]","sigma")) +
  ggtitle("Posterior Density Overlays: Log10 Viral Load (Vague Prior)")

# Convergence summary
conv_sum <- summarise_draws(fit_vl_vague$draws()) %>%
  filter(grepl("beta|sigma", variable)) %>%
  select(variable, mean, sd, rhat, ess_bulk, ess_tail)

kable(conv_sum, caption = "Convergence Diagnostics: Log10 Viral Load (Vague Prior)",
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Posterior Inference Results (Table A4)

```{r posterior-table}
post_tab <- bind_rows(bayes_results)
rownames(post_tab) <- NULL

kable(post_tab,
      caption = "Table A4: Bayesian Posterior Summaries for Hard Drug Use Coefficient",
      col.names = c("Outcome","Prior","Post. Mean","Post. SD",
                    "95% HPDI Low","95% HPDI High","P(|β|>γ | data)"),
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  collapse_rows(columns = 1)
```

## Model Comparison via LOO-IC

```{r loo-comparison}
# Compare models with vs without hard drug variable using LOO-IC
# Fit null models (no hard drug variable) for each outcome
loo_compare_results <- list()

for (o in outcomes) {
  sd_obj <- build_stan_data(o$y, o$bl, prior_sd_hd = 100)

  # Null model: design matrix without hard drug use
  df_cc <- sd_obj$df_cc
  X_null <- model.matrix(~ bl_c + age_c + race_other + bmi_c +
                           educ_hi + smoke_form + smoke_curr, data = df_cc)
  y_vec  <- df_cc[[o$y]]
  K_null <- ncol(X_null)

  null_data <- list(N = nrow(X_null), K = K_null,
                    X = X_null, y = y_vec,
                    beta_prior_mean = rep(0, K_null),
                    beta_prior_sd   = rep(100, K_null),
                    sigma_prior_sd  = 100)

  fit_null <- mod$sample(data = null_data, chains = 4, parallel_chains = 4,
                         iter_sampling = 20000, iter_warmup = 5000,
                         seed = 42, refresh = 0)

  ll_full <- as_draws_matrix(fit_store[[paste0(o$label,"_Vague")]]$draws("log_lik"))
  ll_null <- as_draws_matrix(fit_null$draws("log_lik"))

  loo_full <- loo(ll_full)
  loo_null_res <- loo(ll_null)
  comp     <- loo_compare(loo_null_res, loo_full)

  loo_compare_results[[o$label]] <- data.frame(
    Outcome     = o$label,
    LOO_Full    = round(loo_full$estimates["elpd_loo","Estimate"], 2),
    LOO_Null    = round(loo_null_res$estimates["elpd_loo","Estimate"], 2),
    ΔLOO        = round(diff(comp[,"elpd_diff"]), 2),
    SE_ΔLOO     = round(diff(comp[,"se_diff"]), 2)
  )
}

loo_tab <- bind_rows(loo_compare_results)

kable(loo_tab,
      caption = "Table A5: LOO-IC Model Comparison (Full vs Null Without Hard Drug Use)\n  ΔLOO > 3 favors the full model.",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

---

# Comparison of Frequentist and Bayesian Results

```{r freq-bayes-comparison}
# Compile frequentist results
freq_comp <- freq_results %>%
  rename(Point_Est = Estimate, CI_Low = CI_lower, CI_High = CI_upper) %>%
  mutate(Method = "Frequentist (95% CI)", Prior = "—") %>%
  select(Outcome, Method, Prior, Est = Point_Est, Low = CI_Low, High = CI_High)

# Compile Bayesian vague prior results
bayes_comp <- post_tab %>%
  filter(Prior == "Vague") %>%
  mutate(Method = "Bayesian Vague (95% HPDI)") %>%
  rename(Est = Post_Mean, Low = HPDI_Lower, High = HPDI_Upper) %>%
  select(Outcome, Method, Prior, Est, Low, High)

# Compile Bayesian skeptical prior results
bayes_skept <- post_tab %>%
  filter(Prior == "Skeptical") %>%
  mutate(Method = "Bayesian Skeptical (95% HPDI)") %>%
  rename(Est = Post_Mean, Low = HPDI_Lower, High = HPDI_Upper) %>%
  select(Outcome, Method, Prior, Est, Low, High)

comparison_tab <- bind_rows(freq_comp, bayes_comp, bayes_skept)

kable(comparison_tab,
      caption = "Table A6: Comparison of Frequentist and Bayesian Estimates for Hard Drug Use Effect",
      digits = 4, row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  collapse_rows(columns = 1)
```

```{r comparison-plot}
comparison_tab %>%
  mutate(Method = factor(Method,
    levels = c("Frequentist (95% CI)",
               "Bayesian Vague (95% HPDI)",
               "Bayesian Skeptical (95% HPDI)"))) %>%
  ggplot(aes(x = Est, xmin = Low, xmax = High, y = Method, color = Method)) +
  geom_pointrange(size = 0.8, position = position_dodge(0.5)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ Outcome, scales = "free_x", ncol = 2) +
  scale_color_manual(values = c("#2166AC","#D6604D","#4DAC26")) +
  labs(title = "Figure: Comparison of Frequentist and Bayesian Estimates",
       subtitle = "Hard Drug Use Coefficient (adjusted); error bars = 95% CI/HPDI",
       x = "Estimate", y = NULL, color = NULL) +
  theme_bw() +
  theme(legend.position = "bottom", axis.text.y = element_blank())
```

---

# Prior Sensitivity Analysis

```{r prior-sensitivity}
sens_tab <- post_tab %>%
  select(Outcome, Prior, Post_Mean, Post_SD, HPDI_Lower, HPDI_Upper, P_meaningful)

kable(sens_tab,
      caption = "Table A7: Prior Sensitivity — Vague vs Skeptical Priors",
      col.names = c("Outcome","Prior","Post. Mean","Post. SD",
                    "HPDI Lower","HPDI Upper","P(|β|>γ | data)"),
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  collapse_rows(columns = 1)
```

---

# Results Summary

## Descriptive Findings

Baseline characteristics differed between hard drug users (n = `r sum(analytic$hard_drugs_f=="Yes")`)
and non-users (n = `r sum(analytic$hard_drugs_f=="No")`). Notably, hard drug users had lower BMI,
worse physical QoL at baseline, and were predominantly current smokers. Trajectory plots showed
that non-users experienced larger gains in CD4 count over time and greater viral load suppression.

## Frequentist Results

Hard drug use at baseline was associated with significantly lower CD4+ T cell count at Year 2
(β ≈ −168, p < 0.001) and lower physical QoL (β ≈ −3.3, p = 0.014), after adjusting for baseline
values and all pre-specified covariates. There was no significant association with log10 viral load
or mental QoL.

## Role of Adherence

ART adherence at Year 2 differed by baseline drug use group (see adherence table and chi-square
test above). When adherence was added to the regression models, the hard drug use coefficient for
CD4 count and physical QoL changed by the percentages shown in the mediation table. Outcomes
for which adherence explains >10% of the hard drug association suggest that part of the effect
operates through lower treatment adherence among drug users — a finding with direct clinical relevance.

## Bayesian Results

Under vague priors, Bayesian posterior means closely mirrored frequentist point estimates, as
expected. The skeptical prior attenuated the hard drug use estimate toward zero, particularly for CD4
count and physical QoL, where the data signal is strongest. The posterior probability of a clinically
meaningful effect P(|β₁| > γ | data) was highest for CD4 count.

LOO-IC comparisons indicated that models including hard drug use provided better predictive fit for
CD4 count and physical QoL (ΔLOO > 3), supporting the frequentist findings.

## Comparison of Approaches

Frequentist and Bayesian (vague prior) estimates were highly concordant across all four outcomes.
The skeptical prior modestly reduced point estimates and widened uncertainty intervals, but
conclusions remained qualitatively similar for CD4 and physical QoL, demonstrating robustness
to the choice of prior.

---

# Software

All analyses were conducted in R (version 4.4.1). Key packages: `cmdstanr` for Bayesian modeling
via Stan, `loo` for LOO-IC model comparison, `bayesplot` for MCMC diagnostics, `posterior` for
posterior summaries, `HDInterval` for HPDIs, `tableone` for descriptive tables, and `tidyverse` /
`ggplot2` / `patchwork` for data manipulation and visualization.

```{r session-info, echo=FALSE}
sessionInfo()
```
